---
title: "Splines in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is a Spline?

Wikipedia:   <https://en.wikipedia.org/wiki/Flat_spline>

#### examples of splines in physical use 

<https://duckduckgo.com/?t=ffab&q=spline+boatbuilding&iax=images&ia=images>

#### Mathematics of splines
<https://en.wikipedia.org/wiki/Spline_(mathematics)>

<https://towardsdatascience.com/unraveling-spline-regression-in-r-937626bc3d96>

#### Splines in R    Packages:  splines, splines2

*  B-splines  
*  M-splines  
*  I-splines  
*  convex splines (C-splines)  
[Wikipedia]
<https://www.rdocumentation.org/packages/splines2/versions/0.4.5>


# Let's do this ourselves

#### First create some data

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(splines)
library(splines2)
library(cowplot)

dat <- tibble(x = c(0.0, 2, 5, 10, 20,25, 30,40, 60,80),
              y = c(0.0, 0.05,0.1, 0.3, 5, 10, 20,30, 32, 33))
plot(dat)

```

#### fit a spline function using **knots**
```{r}
# spline degree
degree = 3
# knots
knots <- c(10, 30)

bSpline(dat$x, knots = knots, degr=degree)
```
#### fit a spline function using **degrees of freedom**
```{r}
# degrees of freedom
df = 3

# spline degree
degr = 3         # cubic spline

bSpline(dat$x, df=df, degr=degr)
```


## curve fitting using splines

```{r}
fitmodel <- function(df, degr, dat){
   #main curve fit
   fit <- lm(y ~ bSpline(x, df=df, degr=degr), data=dat)
   fit
}


fit <- fitmodel(df=4, degr = 3, dat)  
summary(fit)
```


```{r}
dat$pred <- fit$fitted.values  
dat$resid <- fit$residuals  
dat 
```

# plotit
```{r}
plotit <- function(dat){
   ggplot(dat) +
      geom_line(aes(x, pred), colour = "gray50") +
      geom_col(aes(x, resid), colour = "red", width=0.3) +
      geom_hline(aes(yintercept=0), colour="brown") +
      geom_point(aes(x, y), colour = "blue", size = 3) 
   
}

fit <- fitmodel(df=4, degr = 3, dat)  
p <- plotit(dat)
p
```


## first derivative
```{r}
# 1st derivative
first.deriv <- lm(y ~ bSpline(x, df=df, degr=degr, derivs=1L), data=dat)

dat$firstD <- first.deriv$fitted.values

p <- p + geom_line(data = dat, aes(x, firstD), color = "green")
p
```

## second derivative
```{r}
# 1st derivative
second.deriv <- lm(y ~ bSpline(x, df=df, degr=degr, derivs=2L), data=dat)

# dat$secondD <- second.deriv$fitted.values

# p <- p + geom_line(data = dat, aes(x, secondD), color = "darkgreen")
# p
```

